---
date: "2019-05-05T00:00:00+01:00"
draft: false
linktitle: R and SNA
menu:
  example:
    name: 1. Basics
    weight: 2
summary: Basics in R adn RStudio
title: 1.Basics
type: docs
weight: 2
---

Follow along with the [power point presentation](https://github.com/EKuki/website/blob/master/content/courses/example/1.Basics_SNA%20using%20R_18March19.pptx) corresponding to this `1. Basic` section.

_**Subsection 1.1**_

[This video](https://youtu.be/b7R8Q04jdrU) provides an intro to R and RStudio that that I am skiping here, so if you are really new to R/RStudio, I would recommend you to watch it. In it, I go along this `1.1` subsection of `1. Basic`.

## How to comment and run code

To write a comment, type `#` before the sentence you want to write.
Anything to the right of a `#` in a script will be ignored by R.

> To comment on a whole paragraph, select the lines you want to comment and press at the same time on your keyboard `Ctrl` + `Shift` + `c`.

> To run: Click anywhere on the line you want to run and either 1) Click `Ctr` + `Enter` or 2) Click on `run` bottom (up right corner of Source/Script)

> If you want to run several lines, highlight them all and do as explained above.

You can get output in the Console simply by typing math in the script and running it:
```{r, echo = TRUE}
3 + 5 # Result is 8
12/2 # Result is 6 
```

##  Assign values to objects
`<-` assigns values on the right to objects on the left.
> PC: Type `Alt` and `-` at the same time. 
> Mac: `Option` and `-`

```{r, echo = TRUE}
weight_kg <- 55    # doesn't print anything, but it has created an object. You can see it in the Environment window, in RStudio.
(weight_kg <- 55)  # Putting parenthesis around the call prints the value of `weight_kg`
weight_kg          # and so does typing (or "calling") the name of the object
```

> Note that R iS cAsE SeNsItIvE!! Try typing `weight_KG`. You will see an error like this one `Error: object 'weight_KG' not found`. Note the capitalizatio of `_KG`.

You can override an object. Be careful!
```{r, echo = TRUE}
weight_kg <- 60 
weight_kg # If you call wight_kg you see that its value is now 60. You can also see this change in the Environment window.
```

**Exercise 1**: Create an object (a vector) that is half the value of weight_kg (use weight_kg <- 60) called "half_weight_kg"

```{r, echo = TRUE}
half_weight_kg <- weight_kg / 2
half_weight_kg
```

You can give whichever name you want to an object, but it is a good idea to choose something that makes sense. For example:

```{r, echo = TRUE}
beautiful_day <- weight_kg / 2
beautiful_day # A bit nonesense...
half_weight_kg
```
> End of Exercise 1

_**Subsection 1.2**_

In [this video](https://youtu.be/l7jsiDXV4yI) I go along this `1.2` subsection of `1. Basic`.
Remember to follow with the power point corresponding to this `1. Basic` section!

To "concatenate" two vectors (aka, to put them together) we use `c()`:
```{r, echo = TRUE}
both_weights<-c(weight_kg, half_weight_kg)
both_weights
```

Another example:
```{r, echo = TRUE}
mum <- "Claire"
dad<- "Pablo"
parents <- c(mum, dad)
```

**Exercise 2**: Create a vector called "daughter" with a female's name and another one called "dog" with a dog's name.

```{r, echo = TRUE}
daughter <- "Anita"
dog <- "Poppy"
```

Create a new vector called "family" with the names of all family members (use this order mum, dad, daugther, dog). 
> With R, we usually can get to the same outcome coding in different ways. My solution is usually one of many.

```{r, echo = TRUE}
family <- c(parents, daughter, dog)
family
class(family)
c(mum, dad, daughter, dog)

family_char <- c("Claire", "Pablo" , "Anita",  "Poppy")
class(family_char)
```
> End Exercise 2

## a.Vectors
```{r, echo = TRUE}
a <- c(1,2,5.3,6,-2,4) # numeric vector
b <- c("one","two","three") # character vector
ce <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
```

### Indexing a `vector`
```{r, echo = TRUE}
ce[2] # Extract second element of `ce`
a[3] # Extract third element of a
a[1:3] # Extrat all elements between the first and the third elements, of `a`
b[c(1,3)] # Extract first and third element of `b`
```

> Note that I did not use the letter `c` to name my vector: I used `ce` instead. The choice of c as the name of a vector is a pretty bad one, as there is a command/function called `c()`! See `?c` or `help(c)`

**Exercise 3**: Extract the first element of the vector family: the `mum`.
```{r, echo = TRUE}
family[1]
```

Extract the forth element of the vector family: the `dog`.
```{r, echo = TRUE}
family[4]
```

Extract both the name of the `dad` and the `daughter` (do it at the same time -using only one command-, if possible).
```{r, echo = TRUE}
family[c(2,3)]
```
> End of Exercise 3


## b.Data frames
```{r, echo = TRUE}
Identification <- c(1,2,3,4)
Colors <- c("red", "white", "red", NA) # Note NA
Succeeded <- c(TRUE,TRUE,TRUE,FALSE)
mydata <- data.frame(Identification, Colors, Succeeded) # Join those vectors together to create a `data.frame`
mydata
names(mydata) <- c("ID","Color","Passed") # Change the variable/column names
```

### Indexing a `data.frame`
```{r, echo = TRUE}
mydata[ ,c(1:2)] # columns 1 and 2 of `mydata` data frame. All rows
```

Other ways to identify observations in a `data.frame`.
```{r, echo = TRUE}
mydata[,c("ID","Color")] # columns ID and Color from `mydata`
mydata$Passed # variable ID in the data frame
```
> The $ operator is used constantly. It retrieves/extracts a variable in a dataset.

**Exercise 4**: Create a data frame called `family_df` with two columns: `name` and `age`. For column `name`, use the previous `family` vector. For column `age`, make up ages for our family
```{r, echo = TRUE}
family
age <- c(45, 48, 10, 7)
family_df <- data.frame(family, age)
family_df
names(family_df) <- c("name","age") # variable names 
family_df
```
> End of Exercise 4

## c.Matrices
```{r, echo = TRUE}
nb_matrix<-matrix(data = 1:20, nrow=5,ncol=4) # The function `matrix()` generates 5 x 4 numeric matrix called `nb_matrix`
nb_matrix

help(matrix) # look for help of the function `matrix()`
?matrix
??matrix
?graph_from_data_frame
??graph_from_data_frame
```

Another example
```{r, echo = TRUE}
cells <- c(1,26,24,68)
mymatrix <- matrix(data = cells, nrow=2, ncol=2, byrow=TRUE)
mymatrix
```

### Indexing a `matrix` (as we did with `data.frames`)
```{r, echo = TRUE}
nb_matrix[,4] # 4th column of nb_matrix
```

**Exercise 5**: Extract the 3rd row of `nb_matrix` 
```{r, echo = TRUE}
nb_matrix[3,]
```

Extract rows 2,3,4 of columns 1,2,3 of `nb_matrix` 
```{r, echo = TRUE}
nb_matrix[2:4,1:3]
nb_matrix[c(2,3,4), c(1,2,3)]
```
> End of Exercise 5


_**Subsection 1.3**_

In [this video](https://youtu.be/bxCB1xXBYcA) I go along this `1.3` subsection of `1. Basic`.
Remember to follow with the power point corresponding to this `1. Basic` section!
